---
title: "Can tt_test() handle empty quadrats?"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = FALSE)
```

```{r}
library(tidyverse)
library(fgeo)
```

Jennifer Baltzer reported that `tt_test()` fails if the census data contains quadrats with no species. Here I run `tt_test()` with the census data that Jennifer sent me, and with a synthetic habitat dataset. I get no error, suggesting that the problem was not in the census data but in the habitat data. I'll need the habitat data to explore what the problem might be.

```{r}
path <- here::here("inst/issues/43/private/scotty.tree1.rdata")
load(path)
scotty <- scotty.tree1
as_tibble(scotty)
```

Show gaps in the census dataset.

```{r, cache=TRUE}
ggplot(scotty, aes(gx, gy)) + 
  geom_point()
```

Create habitat data.

```{r}
# Create elevation data
xy <- expand.grid(x = seq(0, 125, by = 20), y = seq(0, 800, by = 20))
elev <- sample.int(500, nrow(xy), replace = TRUE)
elevation <- bind_cols(tibble::as.tibble(xy), elev = elev)

# Create habitat data from elevation data
hab <- fgeo_habitat(elevation, xdim = 800, ydim = 125, gridsize = 20, n = 4)
hab
```

Run `tt_test()` with all species.

```{r}
all_species <- unique(scotty$sp)
tt_list <- tt_test(scotty,all_species, habitat = hab)
tt_list[1:3]
```

I get no error.
