---
title: "Habitat-species associations"
subtitle: "Given a number of species and habitat types in a plot, is each species significantly aggregated within habitats?"
author: "Mauro Lepore"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Habitat-species associations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
options(digits = 3)

knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  cache = FALSE,
  out.width = "70%",
  fig.align = "center",
  fig.height = 9,
  fig.width = 6,
  fig.asp = 0.618,  # 1 / phi
  fig.show = "hold"
)
```

This article shows how to determine habitat-species associations with the function [`tt_test()`](https://forestgeo.github.io/fgeo.habitat/reference/tt_test), developed by Sabrina Russo, Daniel Zuleta, Matteo Detto, and Kyle Harms. 

To suggest changes to the source code please file a pull request on the GitHub repository of the relevant package. You can search all functions and packages [here](https://forestgeo.github.io/fgeo/articles/fgeo.html#functions). Each function's help file links to its source code.

## Setup

First, install and load ("open") the relevant packages. (If this code chunks isn't clear enough, see [this article](https://fgeo.netlify.com/2018/02/05/2018-02-05-installing-packages-from-github/) for details on how to install packages from GitHub.)

```R
# install.packages("remotes")
remotes::install_github("forestgeo/fgeo.habitat")
```

```{r}
library(dplyr)
library(ggplot2)
library(fgeo.habitat)
library(fgeo.tool)
```

```{r}
# For reproducible results
set.seed(1014)
```

## Load census and habitat data

we will use example datasets that come with __fgeo.habitat__.

```{r}
census <- luquillo_tree6_random

str(census)
```

```{r}
# Creating habitat data from elevation data
habitat <- fgeo.tool::fgeo_habitat(luquillo_elevation, gridsize = 20, n = 4)
str(habitat)
```

To load your own data, you may run something like this:

```R
load("PATH/CENSUS_DATA.rdata")
census <- CENSUS_DATA

load("PATH/HABITAT-DATA.rdata")
habitat <- HABITAT-DATA
```

## Pick data

We will pick alive trees, of 10 mm or more, and of sufficiently abundant species.

```{r}
pick <- filter(
  census,
  # Keep only alive
  status == "A", 
  # Keep dbh of 10 mm or more (drops missing dbh)
  dbh >= 10
)
# Count number of rows per species
pick <- add_count(pick, sp)
# Keep sufficiently abundant trees
pick <- filter(pick, n > 50)

# Summary
unique(select(pick, sp, n))
```

## Overview

Before testing, we can overview the relationship between species an habitats with a plot.

```{r}
# Tweaks
offset <- 20 / 2
habitat2 <- mutate(
  habitat, 
  # Center species and habitat data
  x = gx + offset, 
  y = gy + offset,
  # From continuous to categorical
  habitats = as.factor(habitats)
)
```

```{r, out.width="90%", fig.widh=(6 * 0.9 / 0.7)}
ggplot(pick, aes(x = gx, y = gy)) +
  geom_raster(data = habitat2, aes(x, y, fill = habitats)) + 
  geom_point() +
  coord_fixed() +
  facet_wrap(~sp) +
  labs(fill = "Habitat")
```

## `tt_test()` and any number of species

```{r}
all_sp <- unique(pick$sp)
out <- tt_test(census, all_sp, habitat)
head(out)
```

You can gather all results into a single dataframe -- which lets you use a wide range of tools for data manipulation and visualization.

```{r}
tt_df <- to_df(out)
head(tt_df)

tail(tt_df)
```

## Data science workflows

This final section shows some examples of how you can benefit from storing your results in a dataframe. Compared to a matrix, a dataframe fits better in common workflows for data manipulation and visualization. The dataframe is the most important data structure used in __dplyr__, __ggplot2__, and many other packages. Here are some examples of what you can do with our dataframe output.

(The next few code chunks use [the pipe operator (`%>%`)](http://style.tidyverse.org/pipes.html) to avoid saving intermediary results and to make our code more expressive -- where each line is an imperative statement that communicates our intention.)

* Filtering interesting rows and arranging them as required.

```{r}
tt_df %>% 
  filter(distribution == "aggregated")

tt_df %>% 
  filter(distribution == "neutral", probability > 0.5) %>% 
  arrange(habitat, desc(probability))
```

* Producing analytic summaries.

```{r}
tt_df %>% 
  group_by(sp, distribution) %>% 
  summarise(total_stem_count = sum(stem_count))
```

* Producing visual summaries.

(Unfortunately, __ggplot2__ was developed before the pipe was invented. In ggplot2, we compose functions not with `%>%` but with `+`.)

```{r, out.width="100%", fig.widh=(6 * 0.9 / 0.7)}
ggplot(tt_df, aes(sp, probability)) + 
  geom_col(aes(fill = distribution), position = "dodge") +
  geom_label(aes(label = stem_count)) +
  facet_wrap(vars(habitat)) +
  labs(
    title = "Probability of species distribution by habitat",
    y = "species", 
    caption = "The number on each bar indicates the count of stems."
  ) +
  coord_flip()
```

## `torusonesp.all()` and one species

Before we used `tt_test()`, a wrapper of `torusonesp.all()`. This section shows how you may use `torusonesp.all()` directly to test a single species. The following section shows how to extend the test to any number of species.

See `?torusonesp.all()`; you will find out that this function needs the name of one species, a census, and habitat data. All of these we already have (above). But this function also needs plot dimensions, grid size and the abundance (count) of each species per quadrat-index.

We should find and store the plot dimensions and gridsize because we will need them in more than one function and they must be consistent. To understand which values to choose we may inspect the structure of the habitat ([plot details](https://forestgeo.si.edu/sites/north-america/luquillo)).

```{r}
lapply(habitat, range)
```

You can find information about each ForestGEO site at https://forestgeo.si.edu/sites-all. Also, consider these functions:

```{r}
plot_dimensions <- extract_plotdim(habitat)
plot_dimensions

grid_size <- extract_gridsize(habitat)
grid_size
```

To calculate abundance per quadrat-index you may use the function [`abundanceperquad()`](https://s.si.edu/2JtN94p) from the original CTFSRPackage, or a more conservative wrapper, `abund_index()`, from __fgeo.habitat__.

```{r}
# For details see ?abund_index()
abundance <- abund_index(pick, plot_dimensions, grid_size)

# Show only a few lines of output
str(abundance, list.len = 5)
```

Now that we have all we need, we can run `torusonesp.all()`.

```{r}
one_species <- unique(pick$sp)[[1]]
one_species

out_one <- torusonesp.all(
  one_species, habitat, abundance, plot_dimensions, grid_size
)
out_one
```

## `torusonesp.all()` and any number of species

To iterate over multiple species you may write a for loop, or use a functional such as `lapply()`.

```{r}
all_species <- unique(pick$sp)
lapply(all_species, torusonesp.all, habitat, abundance, plot_dimensions, grid_size)
```
