---
title: "Restructure tt_test() output for easier manipulation"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE)
```

This document explores a new structure for the output of `tt_test()` for easier manipulation and visualization.

Setup.

```{r}
library(fgeo)
```

Original output.
 
```{r}
census <- filter(luquillo_top3_sp, status == "A", dbh >= 10)
habitat <- luquillo_habitat
species <- c("CASARB", "PREMON", "SLOBER")

tt_df <- to_df(tt_test(census, species, habitat))
tt_df
```

This output is easy to visualize:

```{r plot-method, echo=FALSE}
library(ggplot2)

tt_df <- structure(tt_df, class = c("tt_df", class(tt_df)))

plot.tt_df <- function(x, ...) {
  ggplot(x, aes(sp, probability)) + 
    geom_col(aes(fill = distribution), position = "dodge") +
    geom_label(aes(label = stem_count)) +
    facet_wrap(vars(habitat)) +
    labs(
      title = "Probability of species distribution by habitat",
      x = "species", 
      caption = "The number on each bar indicates the count of stems."
    ) +
    coord_flip()
}
```

```{r plot}
plot(tt_df)
```

This output is also easy to manipulate, for example, to count the number of species in each `distribution`-category by `habitat`:

```{r}
count(tt_df, habitat, distribution)
```

This does the same thing. It is more verbose but it may be eaiser to see what's going on.

```{r}
tt_df %>% 
  group_by(habitat, distribution) %>% 
  summarise(n = n_distinct(sp))
```

### Check

Confirm with Daniel that I'm calculating the probabilities correctly. The comments in the next chunk of code come from the interpretation section of `?tt_test()`.

```{r}
probability <- select(tt_df, Obs.Quantile, distribution, probability)

# Note that to calculate the probability for repelled, it is the value given
filter(probability, distribution != "aggregated")

# ... but to calculate the probability for aggregated, it is 1- the value given
filter(probability, distribution == "aggregated")
```
